Structure Learning for $\VAR(p)$ Models

### 2019-05-02 paper/paper.tex
Work on algorithm proof.

\begin{remark}
  Considering the contrapositive of the first part of this theorem (which is what we actually prove below) is enlightening: it says that if $i$ is not a descendent of $j$ (i.e. $j \not \in \anc{i}$) and that there are no common ancestors (i.e. $\anc{i} \cap \anc{j} = \emptyset$) then $j \not \pwgc i$, or $\linE{x_i(t)}{\H_{t - 1}^{(i, j)}} = \linE{x_i(t)}{\H_{t - 1}^{(i)}}$, which is an intuitively natural result.

  The first part is used in part to prove the second (and stronger) statement, which allows common ancestors, but only if they are ``mediated'' by $i$.  For example, this covers the case of a graph $k \rightarrow i \rightarrow j$, where $\{k\} = \anc{i} \cap \anc{j}$, but clearly $j \not \pwgc i$.

@@

Examples of strongly causal graphs include directed trees (or
forests), and DAGs where each node has at most one parent.
Fortunately, the strongly causal property is inherited by subgraphs:

@@

  In this case, node $1$ is a confounder for nodes $2$ and $3$, but
  $x_3(t) = v_3(t) - v_2(t) + x_2(t)$ and $2 \not \pwgc 3$ (even
  though $x_2(t)$ and $x_3(t)$ are contemporaneously correlated)

@@

    \BlankLine
    \caption{Pairwise Graph Recovery}
    \label{alg:pwgr}
    % \TitleOfAlgo{Pairwise Graph Recovery}
    \Input{Pairwise Granger-causality relations between $n$ processes whose joint Granger-causality relations are known to form a strongly causal DAG $\gcg$.}
    \Output{Edges $E = \{(i, j) \in [n] \times [n]\ |\ i \gc j \}$ of the graph $\gcg$.}
    \Initialize{$S_0 = [n]$  \texttt{\# unprocessed nodes}\\
      $E = \emptyset$  \texttt{\# edges of }$\gcg$\\
      % $P_0 = \emptyset$  \texttt{\# layer by layer driving nodes}\\
      $k = 0$ \texttt{\# a counter used only for notation}}
    \BlankLine
    $W_0 \leftarrow \{(i, j)\ |\ i \pwgc j, j \npwgc i\}$  \texttt{\# potential edges}\\
    \While{$S_k \ne \emptyset$}{
      $P_k \leftarrow \{i \in S_{k - 1}\ |\ \forall s \in S_{k - 1}\ (s, i) \not\in W_k\}$  \texttt{\# parents of current layer}\\
      % $P_k \leftarrow P_{k - 1} \cup P_k$ \texttt{\# Expand parent set}\\
      $S_{k + 1} \leftarrow S_k \setminus P_k$ \texttt{\# remove from } $S$\\

      \;
      $C_k \leftarrow \{j \in S_{k + 1}\ |\ \forall s \in S_{k + 1}\ (s, j) \not\in W_k\}$   \texttt{\# potential children of }$P_k$\\

      \For{$r = 0, \ldots, k$}
      {
        $D_{kr} \leftarrow \{(i, j) \in P_{k - r} \times C_k\ |\ (i, j) \in W_k,\ \text{no } i \rightarrow j \text{ path in } E \}$\\%  \texttt{\# edges incident on }$C_k$\\
        $E \leftarrow E \cup D_{kr}$ \texttt{\# add edges to }$E$\\
        % $W_{k + 1} \leftarrow W_k \setminus D_{kr}$  \texttt{\# remove edges from consideration}\\
      }
      $W_{k + 1} \leftarrow W_k \setminus \big(\bigcup_{r = 0}^k D_{kr}\big)$  \texttt{\# remove edges from consideration}\\

@@

\begin{example}

  That we need to proceed backwards through $P_{k - r}$ can be seen from the example of figure \ref{fig:example_fig3}, where if $D_k = \{(i, j) \in \Big(\bigcup_{r = 0}^k P_{k - r}\Big) \times C_k\ |\ i \pwgc j \}$ above we would infer the false positive edge $2 \rightarrow 3$.  Moreover, the same example shows that if simply $D_k = \{(i, j) \in P_k \times C_k\ |\ i \pwgc j \}$, then the edge $1 \rightarrow 3$ would be missed.

\hl{TODO: Inkscape figure}

  \begin{figure}
    \centering
    \label{fig:example_fig3}
    \caption{Example graph for Algorithm \ref{alg:pwgr}}
    \footnotesize{Black arrows indicate true parent-child
      relations.  Red dotted arrows indicate pairwise causality (due to
      ancestor relations), green dash-dotted arrow indicates
      bidirectional pairwise causality (due to the confounding node
      $1$).  Blue groupings indicate each $P_k$ in Algorithm
      \ref{alg:pwgr}.  Repetition is for aesthetics.}

    \begin{subfigure}[b]{0.45\textwidth}
      \includegraphics[width=\linewidth]{example_algorithm.pdf}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
      \includegraphics[width=\linewidth]{example_algorithm2.pdf}
    \end{subfigure}
  \end{figure}
\end{example}

\begin{lemma}
  \hl{Some lemma on the inner loop.}
\end{lemma}
\begin{proof}
  Let $j \in C_k$ and $i \in \bigcup_{r = 0}^k P_{k - r}$.

  First, suppose that $i \in \pa{j}$ and that there is no
  $i \rightarrow j$ path in $E$.  Then since $i \gc j$ we have
  $i \pwgc j$ by proposition \ref{prop:pwgc_anc} so at least one of
  $\{D_{kr} \}_{r = 0}^k$ contains $(i, j)$ and hence $(i, j)$ will be
  added to $E$ upon the end of the iteration.

  Now, still assuming that $i \in \pa{j}$, suppose instead that there
  is already an $i \rightarrow j$ path in $E$.  Using an induction
  hypothesis that $E$ contains edges if and only if they are in
  $\gcg$, then either $(i, j) \in E$ correctly, or there is more than
  one $i \rightarrow j$ path in $\gcg$, which is not possible since
  $\gcg$ is strongly causal.

  Finally, suppose that $i \not \in \pa{j}$.  If $i \not \pwgc j$ then
  we will correctly leave $(i, j)$ absent from $E$, so suppose that
  $i \pwgc j$.  Then we must have either $i \in \anc{j}$ or
  $\anc{i} \cap \anc{j} \ne \emptyset$; suppose first the former.
  Then, there is some $u$ s.t. $i \pwgc u \pwgc j$, so by the
  construction of \hl{it seems} that both $(i, u)$ and $(u, j)$ should
  have already been added to $E$, so we have an $i \rightarrow j$ path
  in $E$ \hl{is this sound?}  Suppose then the latter case above,
  $u \in \anc{i} \cap \anc{j}$ \hl{try to construct a counterexample
    involving this case}.
\end{proof}

@@

  We will first dispense with some trivialities: if $n = 0$ or $n = 1$
  the algorithm is correct, returning on the first iteration with
  $E = \emptyset$.  Suppose then that $n > 1$.  The set
  $P_0$ is non-empty since $\gcg$ is a DAG and therefore
  there are nodes in $\gcg$ without parents, thence if
  $P_0 = \emptyset$ proposition
  \ref{prop:ancestor_properties} would be contradicted.  Now, if
  $P_0 = [n]$ then the graph necessarily contains no edges
  and we will correctly return $E = \emptyset$.

  Moving on from any trivial cases, notice that
  $\mathcal{E} \subseteq W_0$.  This follows since by Proposition
  \ref{prop:pwgc_anc} we have $j \in \pa{i} \Rightarrow j \pwgc i$ and by
  Corollary \ref{cor:bidirectional_edge} there are no bidirectional
  edges in $\gcg$.

>>>>

\begin{remark}
  Considering the contrapositive of the first part of this theorem (which is what we actually prove below) is enlightening: it says that if $i$ is not a descendent of $j$ (i.e. $j \not \in \anc{i}$) and that there are no common ancestors (i.e. $\anc{i} \cap \anc{j} = \emptyset$) then $j \npwgc i$, or $\linE{x_i(t)}{\H_{t - 1}^{(i, j)}} = \linE{x_i(t)}{\H_{t - 1}^{(i)}}$, which is an intuitively natural result.

  The first part is used in part to prove the second (and stronger) statement, which allows common ancestors, but only if they are ``mediated'' by $i$.  For example, this covers the case of a graph $k \rightarrow i \rightarrow j$, where $\{k\} = \anc{i} \cap \anc{j}$, but clearly $j \npwgc i$.

@@

Examples of strongly causal graphs include directed trees (or
forests), DAGs where each node has at most one parent, and each of the
figures of this paper.  Fortunately, the strongly causal property is
inherited by subgraphs:

@@

  In this case, node $1$ is a confounder for nodes $2$ and $3$, but
  $x_3(t) = v_3(t) - v_2(t) + x_2(t)$ and $2 \npwgc 3$ (even
  though $x_2(t)$ and $x_3(t)$ are contemporaneously correlated)

@@

    \BlankLine
    \caption{Pairwise Graph Recovery}
    \label{alg:pwgr}
    % \TitleOfAlgo{Pairwise Graph Recovery}
    \Input{Pairwise Granger-causality relations between $n$ processes whose joint Granger-causality relations are known to form a strongly causal DAG $\gcg$.}
    \Output{Edges $E = \{(i, j) \in [n] \times [n]\ |\ i \gc j \}$ of the graph $\gcg$.}
    \Initialize{$S_0 = [n]$  \texttt{\# unprocessed nodes}\\
      $E = \emptyset$  \texttt{\# edges of }$\gcg$\\
      % $P_0 = \emptyset$  \texttt{\# layer by layer driving nodes}\\
      $k = 0$ \texttt{\# a counter used only for notation}}
    \BlankLine
    $W \leftarrow \{(i, j)\ |\ i \pwgc j, j \npwgc i\}$  \texttt{\# potential edges}\\
    \While{$S_k \ne \emptyset$}{
      $P_k \leftarrow \{i \in S_k\ |\ \forall s \in S_k\ (s, i) \not\in W\}$  \texttt{\# parents of current layer}\\
      % $P_k \leftarrow P_{k - 1} \cup P_k$ \texttt{\# Expand parent set}\\
      $S_{k + 1} \leftarrow S_k \setminus P_k$ \texttt{\# remove from } $S$\\

      \;
      $C_k \leftarrow \{j \in S_{k + 1}\ |\ \forall s \in S_{k + 1}\ (s, j) \not\in W\}$   \texttt{\# potential children of }$P_k$\\

      \For{$r = 0, \ldots, k$}
      {
        \label{alg:inner_loop}
        $D_{kr} \leftarrow \{(i, j) \in P_{k - r} \times C_k\ |\ (i, j) \in W,\ \text{no } i \rightarrow j \text{ path in } E \}$\\%  \texttt{\# edges incident on }$C_k$\\
        $E \leftarrow E \cup D_{kr}$ \texttt{\# add edges to }$E$\\
        % $W_{k + 1} \leftarrow W_k \setminus D_{kr}$  \texttt{\# remove edges from consideration}\\
      }
      % $W_{k + 1} \leftarrow W_k \setminus \big(\bigcup_{r = 0}^k D_{kr}\big)$  \texttt{\# remove edges from consideration}\\

@@

\begin{example}
  The set $W$ collects all of the edges that may be edges of $\gcg$.
  In reference to figure \ref{fig:example_fig3}, each of the solid
  black edges, as well as the dotted red edges will be included, but
  \textit{not} the bidirectional green dash-dotted edge.  That we exclude
  bidirectional edges is important since it will provide guarantees
  that $P_k$ is non-empty.  The groupings $P_0, \ldots, P_3$
  are also indicated in figure \ref{fig:example_fig3}.

  The algorithm proceeds first with the parentless nodes $1, 2$ on the
  initial iteration where the edge $(1, 3)$ is added to $E$.  On the
  next iteration, the edges $(3, 4), (2, 4)$ are added, and the false
  edge $(1, 4)$ is excluded due to the path
  $1 \rightarrow 3 \rightarrow 4$ in E.  Finally, edges
  $(3, 5), (4, 6)$ are added, and the false
  $(1, 5), (1, 6), (3, 6), (2, 6)$ edges are similarly excluded due to
  the ordering of the inner loop.

  \begin{figure}
    \centering
    \caption{Example graph for Algorithm \ref{alg:pwgr}}
    \footnotesize{Black arrows indicate true parent-child
      relations.  Red dotted arrows indicate pairwise causality (due to
      ancestor relations), green dash-dotted arrow indicates
      bidirectional pairwise causality (due to the confounding node
      $1$).  Blue groupings indicate each $P_k$ in Algorithm
      \ref{alg:pwgr}.  Repetition is for aesthetics.}
    \label{fig:example_fig3}

    \begin{subfigure}[b]{0.45\textwidth}
      \includegraphics[width=\linewidth]{example_algorithm.pdf}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
      \includegraphics[width=\linewidth]{example_algorithm2.pdf}
    \end{subfigure}
  \end{figure}

  That we need to proceed backwards through $P_{k - r}$ as in the
  inner loop of line \ref{alg:inner_loop} can also be seen from this
  example, where if instead
  $D_k = \{(i, j) \in \Big(\bigcup_{r = 0}^k P_{k - r}\Big) \times
  C_k\ |\ i \pwgc j \}$ we would infer the false positive edge
  $1 \rightarrow 4$.  Moreover, the same example shows that if simply
  $D_k = \{(i, j) \in P_k \times C_k\ |\ i \pwgc j \}$, then the edge
  $1 \rightarrow 3$ would be missed.
\end{example}

\begin{lemma}
  \label{lem:inner_loop_lemma}
  At any given iteration, $k$, of Algorithm \ref{alg:pwgr}, if we have
  $e \in E$ if and only if $e \in \mathcal{E}$, then the inner loop (line \ref{alg:inner_loop})
  will add $(i, j) \in \big(\bigcup_{r = 0}^kP_{k - r}\big) \times C_k$ to $E$
  if $(i, j) \in \mathcal{E}$.  i.e. any real edge will be included.
\end{lemma}
\begin{proof}
  Let $j \in C_k$, $i \in \bigcup_{r = 0}^k P_{k - r}$, and assume
  $i \in \pa{j}$.

  First, suppose that there is no $i \rightarrow j$ path in $E$.  Then
  since $i \gc j$ we have $i \pwgc j$ and $j \npwgc i$ (i.e.
  $(i, j) \in W$) by Proposition \ref{prop:pwgc_anc} and Corollary
  \ref{cor:bidirectional_edge} so at least one of
  $\{D_{kr} \}_{r = 0}^k$ contains $(i, j)$ and hence $(i, j)$ will be
  added to $E$ upon the end of the iteration.

  Now, suppose instead that there is already an $i \rightarrow j$ path
  in $E$.  Given our assumption that $E$ contains edges if and only if
  they are edges of $\gcg$, then either we already have $(i, j) \in E$
  correctly, or there is more than one $i \rightarrow j$ path in
  $\gcg$, which is not possible since $\gcg$ is strongly causal.
\end{proof}

\begin{lemma}
  \label{lem:W_subset_E}
  $W \subseteq \mathcal{E}$
\end{lemma}
\begin{proof}
  By proposition \ref{prop:pwgc_anc} we have
  $j \in \pa{i} \Rightarrow j \pwgc i$ and by Corollary \ref{cor:bidirectional_edge}
  there are no bidirectional edges in $\gcg$.
\end{proof}

\hl{Must show that Algorithm terminates in $n$ steps.}
\begin{lemma}
  For any $i \in P_k$ we have $i \in C_{k - 1}$.  \hl{[Can't yet prove
    this part] Moreover, for each $i \in \bigcup_{k = 0}^n P_k$, if
    $(s, i) \in W$ then $(s, i) \in E$ upon the termination of
    Algorithm} \ref{alg:pwgr}.
\end{lemma}
\begin{proof}
  We prove the second part by induction, the first part being
  established along the way.  By construction $P_0$ has no parent
  nodes in $\gcg$ so the base case is immediate.  Fix now some
  $k \ge 1$, $i \in P_k$, and let $s$ be such that $(s, i) \in W$.
  For every $s$ such that $(s, i) \in W$ we have $s \not\in S_k$ since
  otherwise $i \not\in P_k$.  Therefore
  $s \in \bigcup_{r = 0}^{k - 1}P_{k - r}$, which implies
  $s \not\in S_k$, and finally that $i \in C_{k - 1}$.  Then by Lemma
  \ref{lem:W_subset_E} every edge in $\mathcal{E}$ is also present in
  $W$, and by Lemma \ref{lem:inner_loop_lemma} \hl{[We can't conclude
    here because of the strong conditions required on lemma}
    \ref{lem:inner_loop_lemma}]
\end{proof}

%   Finally, suppose that $i \not \in \pa{j}$.  If $i \npwgc j$ then
%   we will correctly leave $(i, j)$ absent from $E$, so suppose that
%   $i \pwgc j$.  Then we must have either $i \in \anc{j}$ or a confounding 
%   $k \in \anc{i} \cap \anc{j}$; suppose first the former.
%   Then, there is some $u$ s.t. $i \pwgc u \pwgc j$, so by the
%   construction of \hl{it seems} that both $(i, u)$ and $(u, j)$ should
%   have already been added to $E$, so we have an $i \rightarrow j$ path
%   in $E$ \hl{is this sound?}  Suppose then the latter case above,
%   $u \in \anc{i} \cap \anc{j}$ \hl{try to construct a counterexample
%     involving this case}.
% \end{proof}

@@

  We will first dispense with some trivialities: if $n = 0$ or $n = 1$
  the algorithm is correct, returning on the first iteration with
  $E = \emptyset$.  Suppose then that $n > 1$.  The set
  $P_0$ is non-empty since $\gcg$ is a DAG and therefore
  there are nodes in $\gcg$ without parents (if
  $P_0 = \emptyset$ proposition
  \ref{prop:ancestor_properties} would be contradicted).  Now, if
  $P_0 = [n]$ then the graph necessarily contains no edges
  and we will correctly return $E = \emptyset$.

  Moving on from any trivial cases, notice that
  $\mathcal{E} \subseteq W$.  This follows since by Proposition
  \ref{prop:pwgc_anc} we have $j \in \pa{i} \Rightarrow j \pwgc i$ and by
  Corollary \ref{cor:bidirectional_edge} there are no bidirectional
  edges in $\gcg$.
